---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# pafr

Read, manipulate and visualize 'Pairwise mApping Format' data in R

## Under contstruction

This package is in the process of being turned from reseach code into a nice
usable package. Much of the code is already here, but documentation tests and
added-flexibility are still being added. 

Nevertheless, you can check out the code and run it for youself. Here's how to
make a basic whole-genome dotplot from a PAF file:

```{r, dotplot}
library(pafr)
test_alignment <- system.file("extdata", "fungi.paf", package="pafr")
ali <- read_paf(test_alignment)
dotplot(ali)
```

## Installation

If you want to play with the code already here, you can install this package
using devtools

```r
#install.packages(pafr)
devtools::install_github("dwinter/pafr")
```

## Bugs/Issues

Please use the issue tracker on this repo to let us know about any bugs or
issues that arise as you use this package
